Noisy Work: A Review of The Uncertainty Visualisation Literature

### 2024-03-01 paper.qmd
Big restructure of paper

# Introduction

@@

# Similar work
> Edit note: I think with the addition of more summary papers I should just summarise common themes of issues the papers ran into instead of detailling the work paper by paper.

@@

# Defining Uncertainty
Uncertainty visualisation is not made any easier by the fact that the term "uncertainty" lacks a commonly accepted definition in the literature. @Spiegelhalter2017 even commented that the definition of uncertainty essentially "depends on who you ask". The disarray of organisation and definitions within the uncertainty literature is a common sentiment expressed by reviewers[@Spiegelhalter2017; @Kinkeldey2014]. This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is. In this section we will establish exactly what we mean when we talk about uncertainty.

## Coloquial Definitions
The most common definition of uncertainty used in papers that discuss visualising it are colloquial. Most definitions make some nod towards variance, precision, trust, error, missing values or another related concepts. Since *uncertainty*, unlike *probability*, does not have a strict definition that is established in mathematics, there is large variety in its usage. Some consider it to be synonymous with probabiltiy, some consider it to be an encompassing umbrella term of which probability is only an element, and some consider it to by something else entirely. This lack of a definition leads to uncertainty being swapped out freely with other vague words such as ambiguity and risk. Definitions presented in papers that discuss uncertainty visualisation want it to be encompassing and include everything a layperson might consider when trying to consider "uncertainty". The issue with these broad definitions is is that in order to visualize something, it needs to be quantified, and in order to quantify something, we need to know exactly what it is we are counting. This discrepancy appears regularly in the uncertainty visualisation literature. Often a vague definition such as "uncertainty is anything that isn't deterministic", or even a refusal to define uncertainty at all in an experiments introduction will be accompanied by a highly specific quantification of uncertainty as a probability density distribution of a forecast when providing the visualisation. Ironically it seems that many of the papers that refer to themselves as "uncertainty" papers are just suffering from a form of uncertainty, a lack of precision.[^1]

@@

## Taxonomy Definitions
### Taxonomies of Uncertainty
> Edit Note: A lot of this is going to be moved to the uncertainty definition specific section of the vis lit review and the point here should just be "this is a mess" instead of trying to find any reason in it. My opinion and goal for this section has changed a lot since writing this (specifically deciding that the encapsulating taxonomy doesnt fix the problem and a maths definition is needed) so it is one of the sections that needs big rewrites.

@@

### Taxonomies of Visual Uncertainty
> Edit Notes: Similar to the "similar works" section, I think this section would be better off with a short sentance about each paper and comments on their similarities because a paragraph for each paper is too much

There are some taxonomies of visual uncertainty. These taxonomies seem to seek to define "uncertainty visualisation" and a combined term rather than as two separate words. We largely disagree with this conceptualization of uncertainty visualisations as it blurs the line between the mathematical estimations and assumptions (the uncertainty part) and the visual depiction of those mathematical objects (the visualisation part). While we will not focus on these taxonomies, it is still important to mention them and highlight the reasons they are not good practice.

## Contextual Information
> Edit Note: This section should probably be moved to the visualisation specific section, although that depends on the tone that I end up with after the rewrites of the earlier sections.

# General Attitude Towards Uncertainty (Uncertainty Communication Boradly)

# Uncertainty Visualisation Literature Summary
The previous sections of this paper established several issues in uncertainty communication at large. In this section, we will discuss the current literature on uncertainty visualisation, how the uncertainty communication issues more broadly manifest in the specific case of visualisation, and what general lessons we can take from the existing work.

## Defining uncertainty with resepct to inference
When we consider mapping uncertainty, a reasonable questions that should come to mind is "uncertainty about what?". Every single thing concept or activity in statistics does not have uncertainty. For example, descriptive statistics cannot have uncertainty as they are just describing statistics as they are, the sample mean is the sample mean, we are not uncertain about its value. A sample variance describes dispersion, which is also how we describe uncertainty, but the sample variance in of itself does not have any uncertainty. Most of the mathematical uncertainty concepts have this idea embedded in them. For example, a confidence interval is an interval drawn around the sample mean that will theoretically contain the *populaton mean* 95% of the time. This issue has been noticed several times, however it is often brought up as a *task* dependence rather than an understanding that uncertainty is inherrently linked to inference on a particular statistic. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failling to do it well can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. Health have a wide range of metrics used to communicate different risks [@@Spiegelhalter2017] indicating they utilise new statistics based off the specific aspect of risk that is relevant to the issue. This should make it clear that uncertainty only exists when we are performing inference. Therefore it only makes sense to describe uncertainty in terms of something to be uncertain *about*, it is not some a latent feature of statistics and data.

@@

## Uncertainty evaluation experiments repeat visualisation studies
The first issue in uncertainty visualisation is there does not seem to be a clear rule for what makes a particular visualisation a depiction of uncertainty. This problem seems to start at the lack of the mathematical definition of uncertainty which then bleeds into a lack of a definition for an "uncertainty visualisation". For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a large number of uncertainty visualisation experiments. How can a chart omit uncertainty according to the grammar of graphics, but be used to describe it in several evaluation studies? The answer to this question is related to the definition problem discussed earlier, but to understand how, we will need to go through the grammar of graphics, and its absence in the uncertainty visualisation literature.

@@

@@

## Measuring uncertainty as noise, not signal
Uncertainty is noise. This may be an obvious statement, but what is not obvious is that the existing literature sees noise as a signal. When we directly ask "What is the varinance of this variable" the variance *becomes* the signal. Noise and signal are not an inherrent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *supresses* it. The replication problem in uncertainty visualisation, where experiments are simply repeating established visualisation work, comes from the misunderstandings established in previous sections. Uncertainty is poorly defined, therefore authors do not understand it is not an inherrent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations. This situation is a bit of a mess. It also highlights a unique and facinating problem faced by uncertainty visualisation. If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?

@@

@@

Authors also occasionally use trust as a proxy for uncertainty. Similar to measuring uncertainty through decision quality, this method also has its own series of issues. If the purpose of displaying uncertainty information is to appropriately hedge a signal with noise, then it should be assumed that trust is only related to uncertainty communication through increased transparency and honesty. It does seem to be the case that uncertainty communication increases trust because it is a proxy for transparency *(add citations)*. Unfortunately a large proportion of the literature discusses trust as something that is directly related to uncertainty visualisation rather than as an observable by product. In viewing uncertainty communication as directly related to trust, not related through the proxy of transparency, several unobserved variables are conflated. The amount of uncertainty that is present in the information, whether or not the information is trustworthy or makes sense, prior beliefs of the participants, and the trustworthiness of the source are a few examples of variables that are conflated when authors directly consider trust to be of direct interest. Considering trust, and not transparency, as the metric of importance in uncertainty communication leads to a questionable subtext that argues against transparency. @Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy".Being concerned about audiences perception of trust, without first establishing if what we are communicating is *trustworthy*, leads to the implication that details that result in information not being considered trustworthy should be avoided. Authors that imply that a decrease in trust or an increase in ambiguity or risk aversion are metrics of importance in of themselves do not understand the purpose of uncertainty communication as methods to increase transparency. Science communication should be primarily concerned with accuracy, setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication. The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature [@Zhao2023].

@@

@@

It is unclear how participants are supposed to incorperate uncertainty information into their responses when uncertainty visualisation authors themselves are unsure. Other authors interpret this difficulty as a combination of the uncertainty and task complexity [@Kinkeldey2014], and not as a general confusion. All these examples sever to illustrate that *we dont know how to measure uncertainty* and if we don't know how to measure it, *we cannot evaluate the quality of a visualisation*. It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis, but this idea also has a miriad of issues that come with it.  There is already a series of issues with the reject/do not reject concepts in hypothesis testing, so it is already questionable to set the results of a hypothesis test as the ground truth in an uncertainty visualisation experiment, however these may be additional problems that go beyond the binary nature of the issue. @Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests. This increased sensisitivity could be the result of another aspect of uncertainty that is ignored. Statstical tests are typically built upon a series of assumptions, and it is difficult to identify *which* assumption failure caused the data to fail the statistical tests. Lineup plots allow us to see *how* our data is different to the assumed distribution, and better understand *why* our data may have failled a test ^[Did Patricks paper compare statistical tests to lineup protocols in the event an earlier assumption of the tests, such as, independent observation? I think he did check some. I wonder if you can quantify the information difference you get from a rejected statistical test vs a rejected line plot]. These experiments utilised the lineup protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury]. This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same. @fig-hypyvs shows the conceptual difference between the lineup protocol and uncertainty visualisation. A lineup protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significiantly difference) while an uncertainty plot displays the variance of the estimated value and assesses if the null of "no signal" is within this plot (e.g. if the error bars overlap with zero). ^[Does this flip have mathematical implications in how the plots should be assessed?] This implies a visible signal in an uncertainty plot should indicate some divergence from the null.

>>>>

# Background

## Introduction

@@

## Similar work
> Edit note: I think with the addition of more summary papers I should just summarise common themes of issues the papers ran into instead of detailling the work paper by paper.

@@

# The Uncertainty Definition Problem
## Definitions in Existing Work
Uncertainty visualisation is not made any easier by the fact that the term "uncertainty" lacks a commonly accepted definition in the literature. @Spiegelhalter2017 even commented that the definition of uncertainty essentially "depends on who you ask". The disarray of organisation and definitions within the uncertainty literature is a common sentiment expressed by reviewers[@Spiegelhalter2017; @Kinkeldey2014]. This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is. In this section we will establish exactly what we mean when we talk about uncertainty.

The most common definition of uncertainty used in papers that discuss visualising it are colloquial. Most definitions make some nod towards variance, precision, trust, error, missing values or another related concepts. Since *uncertainty*, unlike *probability*, does not have a strict definition that is established in mathematics, there is large variety in its usage. Some consider it to be synonymous with probabiltiy, some consider it to be an encompassing umbrella term of which probability is only an element, and some consider it to by something else entirely. This lack of a definition leads to uncertainty being swapped out freely with other vague words such as ambiguity and risk. Definitions presented in papers that discuss uncertainty visualisation want it to be encompassing and include everything a layperson might consider when trying to consider "uncertainty". The issue with these broad definitions is is that in order to visualize something, it needs to be quantified, and in order to quantify something, we need to know exactly what it is we are counting. This discrepancy appears regularly in the uncertainty visualisation literature. Often a vague definition such as "uncertainty is anything that isn't deterministic", or even a refusal to define uncertainty at all in an experiments introduction will be accompanied by a highly specific quantification of uncertainty as a probability density distribution of a forecast when providing the visualisation. Ironically it seems that many of the papers that refer to themselves as "uncertainty" papers are just suffering from a form of uncertainty, a lack of precision.[^1]

@@

### Taxonomy Definitions
> Edit Note: A lot of this is going to be moved to the uncertainty definition specific section of the vis lit review and the point here should just be "this is a mess" instead of trying to find any reason in it. My opinion and goal for this section has changed a lot since writing this (specifically deciding that the encapsulating taxonomy doesnt fix the problem and a maths definition is needed) so it is one of the sections that needs big rewrites.

@@

## The role of inference
When we consider mapping uncertainty, a reasonable questions that should come to mind is "uncertainty about what?". Every single thing concept or activity in statistics does not have uncertainty. For example, descriptive statistics cannot have uncertainty as they are just describing statistics as they are, the sample mean is the sample mean, we are not uncertain about its value. A sample variance describes dispersion, which is also how we describe uncertainty, but the sample variance in of itself does not have any uncertainty. Most of the mathematical uncertainty concepts have this idea embedded in them. For example, a confidence interval is an interval drawn around the sample mean that will theoretically contain the *populaton mean* 95% of the time. This issue has been noticed several times, however it is often brought up as a *task* dependence rather than an understanding that uncertainty is inherrently linked to inference on a particular statistic. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failling to do it well can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. Health have a wide range of metrics used to communicate different risks [@Spiegelhalter2017] indicating they utilise new statistics based off the specific aspect of risk that is relevant to the issue. This should make it clear that uncertainty only exists when we are performing inference. Therefore it only makes sense to describe uncertainty in terms of something to be uncertain *about*, it is not some a latent feature of statistics and data.

@@

# The Uncertainty Visualisation Definition Problem

> Edit Notes: Similar to the "similar works" section, I think this section would be better off with a short sentance about each paper and comments on their similarities because a paragraph for each paper is too much

## Uncertainty Visualisations in Existing work

### Taxonomies of Uncertainty Visualisations
There are some taxonomies of visual uncertainty. These taxonomies seem to seek to define "uncertainty visualisation" and a combined term rather than as two separate words. We largely disagree with this conceptualization of uncertainty visualisations as it blurs the line between the mathematical estimations and assumptions (the uncertainty part) and the visual depiction of those mathematical objects (the visualisation part). While we will not focus on these taxonomies, it is still important to mention them and highlight the reasons they are not good practice.

The first issue in uncertainty visualisation is there does not seem to be a clear rule for what makes a particular visualisation a depiction of uncertainty. This problem seems to start at the lack of the mathematical definition of uncertainty which then bleeds into a lack of a definition for an "uncertainty visualisation". For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a large number of uncertainty visualisation experiments. How can a chart omit uncertainty according to the grammar of graphics, but be used to describe it in several evaluation studies? The answer to this question is related to the definition problem discussed earlier, but to understand how, we will need to go through the grammar of graphics, and its absence in the uncertainty visualisation literature.

### Contextual Information
> Edit Note: This section should probably be moved to the visualisation specific section, although that depends on the tone that I end up with after the rewrites of the earlier sections.

## "Uncertainty Visualisation" is not defined

@@

### Uncertainty as signal

@@

# Measureing noise, not signal

## Uncertainty vs Risk
Uncertainty is noise. This may be an obvious statement, but what is not obvious is that the existing literature sees noise as a signal. When we directly ask "What is the varinance of this variable" the variance *becomes* the signal. Noise and signal are not an inherrent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *supresses* it. The replication problem in uncertainty visualisation, where experiments are simply repeating established visualisation work, comes from the misunderstandings established in previous sections. Uncertainty is poorly defined, therefore authors do not understand it is not an inherrent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations. This situation is a bit of a mess. It also highlights a unique and facinating problem faced by uncertainty visualisation. If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?

@@

The previous sections of this paper established several issues in uncertainty communication at large. In this section, we will discuss the current literature on uncertainty visualisation, how the uncertainty communication issues more broadly manifest in the specific case of visualisation, and what general lessons we can take from the existing work.

## Decision Making

@@

## Trust
Authors also occasionally use trust as a proxy for uncertainty. Similar to measuring uncertainty through decision quality, this method also has its own series of issues. If the purpose of displaying uncertainty information is to appropriately hedge a signal with noise, then it should be assumed that trust is only related to uncertainty communication through increased transparency and honesty. It does seem to be the case that uncertainty communication increases trust because it is a proxy for transparency *(add citations)*. Unfortunately a large proportion of the literature discusses trust as something that is directly related to uncertainty visualisation rather than as an observable by product. In viewing uncertainty communication as directly related to trust, not related through the proxy of transparency, several unobserved variables are conflated. The amount of uncertainty that is present in the information, whether or not the information is trustworthy or makes sense, prior beliefs of the participants, and the trustworthiness of the source are a few examples of variables that are conflated when authors directly consider trust to be of direct interest. Considering trust, and not transparency, as the metric of importance in uncertainty communication leads to a questionable subtext that argues against transparency. @Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy".Being concerned about audiences perception of trust, without first establishing if what we are communicating is *trustworthy*, leads to the implication that details that result in information not being considered trustworthy should be avoided. Authors that imply that a decrease in trust or an increase in ambiguity or risk aversion are metrics of importance in of themselves do not understand the purpose of uncertainty communication as methods to increase transparency. Science communication should be primarily concerned with accuracy, setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication. The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature [@Zhao2023].

@@

## Other attemps to capture uncertainty

@@

## Uncertainty is not hypothesis testing
It is unclear how participants are supposed to incorperate uncertainty information into their responses when uncertainty visualisation authors themselves are unsure. Other authors interpret this difficulty as a combination of the uncertainty and task complexity [@Kinkeldey2014], and not as a general confusion. All these examples sever to illustrate that *we dont know how to measure uncertainty* and if we don't know how to measure it, *we cannot evaluate the quality of a visualisation*. It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis, but this idea also has a miriad of issues that come with it.  There is already a series of issues with the reject/do not reject concepts in hypothesis testing, so it is already questionable to set the results of a hypothesis test as the ground truth in an uncertainty visualisation experiment, however these may be additional problems that go beyond the binary nature of the issue. @Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests. This increased sensisitivity could be the result of another aspect of uncertainty that is ignored. Statstical tests are typically built upon a series of assumptions, and it is difficult to identify *which* assumption failure caused the data to fail the statistical tests. Lineup plots allow us to see *how* our data is different to the assumed distribution, and better understand *why* our data may have failled a test ^[Did Patricks paper compare statistical tests to lineup protocols in the event an earlier assumption of the tests, such as, independent observation? I think he did check some. I wonder if you can quantify the information difference you get from a rejected statistical test vs a rejected line plot]. These experiments utilised the lineup protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury]. This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same. @fig-hypyvs shows the conceptual difference between the lineup protocol and uncertainty visualisation. A lineup protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significiantly difference) while an uncertainty plot displays the variance of the estimated value and assesses if the null of "no signal" is within this plot (e.g. if the error bars overlap with zero). ^[Does this flip have mathematical implications in how the plots should be assessed?] This implies a visible signal in an uncertainty plot should indicate some divergence from the null.
